---
title: "Assignment 12 - Connectivity"
output: html_notebook
---


```{r}
rm(list=ls())

require(tidyverse)
require(terra)
require(sf)
require(gdistance)
require(igraph)
require(fitdistrplus)
require(fdrtool)
require(raster)
#require(plot)
```

```{r}

# Read in the landcover raster
land = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week10/panther_landcover.tif')

# Identify the crs of the landcover raster
crs(land)

#Look at th raster's resolution
res(land)



# Bring in the public lands shapefile
public = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week12/panther_publicland.shp')

# Set the crs of the public lands shapefile
st_crs(public) = crs(land)

# And look at it
data.frame(public)
```

```{r, warning=F}
publicCentroids = st_centroid(public)

plot(land)
plot(st_geometry(public), add=T)
plot(st_geometry(publicCentroids), add=T, col='red')
```

```{r}
classification = read.table('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/resistance%20reclass.txt', header=T)
landCost = classify(land, classification[,c(1,3)])
landCost = raster(landCost)

plot(landCost)
plot(st_geometry(public), add=T)
plot(st_geometry(publicCentroids), add=T, col='red')
```

# Challenge 1 (5 points)

In the lab, we calculated Euclidean distance, least-cost distance, commute distance (based on circuit theory), and a randomized shortest path distance. Examine the correlation among these 4 metrics of effective distance. Which tend to be the longest, which tend to be the shortest, and why? In what situations might you expect to find strong differences between Euclidean distance and the other metrics of effective distance?

```{r}
#Euclidean distance 
geoDist = pointDistance(publicCentroids, lonlat=F)
geoDist = as.dist(geoDist)
geoDist

#Least-cost distance
landCond = transition(1/landCost, transitionFunction = mean, 8)
landCond = geoCorrection(landCond, type='c', multpl=F)

lcDist = costDistance(landCond, st_coordinates(publicCentroids))
lcDist

#Commute distance
circuitDist = commuteDistance(landCond, st_coordinates(publicCentroids))
circuitDist

#Randomized shortest path distance
rspDist1.2 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[2,], theta=0.001)
rspDist1.2



```
Commute distances tend to be the longest pathing because this function assumes random pathing using circuit theory, which expects random pathings along the areas of lowest resistance regardless of length. Euclidean distances tend to have the shortest pathings because these distances are straight-line pathing without considering any potential landscape resistance. I would expect larger differences between Euclidean distances and other pathings in ecosystems where straight-line pathings have very high resistances. Examples of this are if the straight-line path contains a barrier such as a wall or a landcover type that the study species cannot traverse such as open water for FL panthers.

$\color{red}{\text{I was hoping to see you calculate this for multiple patches and look at correlation, but good work on your reasoning. +3}}$


# Challenge 2 (5 points)

In the lab we developed a 10% least-cost corridor between two protected areas. Identify what proportion of this corridor is comprised of each landcover type based on the land raster. To do this, make sure that both the land and leastCostCorridor10 rasters are spatRasters. Then mask land with leastCostCorridor10. The values in the resulting raster will be representative of the landcover values in the corridor. Summarise this information and refer back to the classification dataframe to link the landcover numbers with landcover types. What are the three most common landcover types within this corridor, and does this make sense given what you know about the resistance to movement within such landcover types?

```{r}
# Choose the extent for cropping
fpwrOssfExtent = extent(642000, 683000, 237000, 298000)

# Crop the landcover layer
landSub = crop(land, fpwrOssfExtent)

# Crop the cost layer and create a transition matrix
landCostSub = crop(landCost, fpwrOssfExtent)
landCondSub = transition(1/landCostSub, transitionFunction=mean, 8)
landCondSub = geoCorrection(landCondSub, type='c', multpl=F)

fpwrCost = accCost(landCondSub, st_coordinates(publicCentroids)[5,])
ossfCost = accCost(landCondSub, st_coordinates(publicCentroids)[3,])

plot(fpwrCost)
plot(ossfCost)

leastCostCorridor = overlay(fpwrCost, ossfCost, fun=function(x, y){return(x+y)})
plot(leastCostCorridor)

#Calculate the 10% quantile for the leastCostCorridor raster
quantile10 = quantile(leastCostCorridor, probs=0.1, na.rm=T)
quantile10

#Create a new raster with a value of 1 if it has a summed cost in the lower 10% and a NA otherwise
leastCostCorridor10 = leastCostCorridor
values(leastCostCorridor10) = NA
leastCostCorridor10[leastCostCorridor < quantile10] = 1

#Plot this raster and look at it
plot(leastCostCorridor10, legend=F, axes=F)
points(publicCentroids, col='grey30') 

```

```{r}
leastCostCorridor10.2 = rast(leastCostCorridor10)
crs(leastCostCorridor10.2) = crs(land)

CorridorComposition = crop(land,leastCostCorridor10.2)
CorridorComposition = mask(CorridorComposition, leastCostCorridor10.2)
table(matrix(CorridorComposition))
```
The most common landcover types are hardwood, hardwood swamp, dry prarie, orchard, and cropland. Both hardwood and hardwood swap share the most landcover type in the least cost corridor with 4 units each. Dry prarie, orchard, and cropland share the second most landcover in the least cost corridor with 3 units each. It makes sense that these landcover types are the most common in the least cost corridor given these landcover types are either preferred habitat types or they are landcover types that would be easier for panthers to navigate due to being drier landcover with vegetation. 

$\color{red}{\text{I'm not exactly sure what went wrong here, but cypress swamp should be at the top.  Either way, good work. +4}}$


# Challenge 3 (5 points)

In the lab, we used the passage() function to conduct a randomized shortest path analysis mapping connectivity among two protected areas. We initially set theta to 0 thus resulting in a circuit theory based map of connectivity. Using the same passage() function, conduct the analysis again using values of theta = 0.0001 and theta = 0.001. Plot the resulting maps. What patterns do you see emerge as theta increases? Comment on the tradeoffs between biological realism and ease of conservation planning based on these three different maps of connectivity.

```{r}
passageMapT0 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0)
plot(passageMapT0)

passageMapT1 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0.0001)
plot(passageMapT1)

passageMapT2 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0.001)
plot(passageMapT2)
```
What patterns do you see emerge as theta increases? Comment on the tradeoffs between biological realism and ease of conservation planning based on these three different maps of connectivity.

As theta increases, we start to see a least-cost corridor emerge. If theta continued to increase we would get a least-cost path. Although organisms do not always or often take the least-cost corridor, organism also do not likely take a completely random path either since if they can see a higher resistent path, they will likely take a visibly observable path with lower resistance (like walking through a valley instead of over a mountain). If the landscape lack much topographic diversity or changes in ecosystem, then a near random movement should be expected (like overlapping magnetic fields). Therefore, if the space and resources are available, conservation planners should try to make loose corridor spaces available for high-resistence landscapes and large-scale protected areas for low-resistence ecosystems because this is the most accurate connectivity of organisms between patches. However, this is usually not an available option to managers due to land use for urban or agriculture. This is why managers develop pathways when building connectivity. 

$\color{red}{\text{Excellent. +5}}$

# Challenge 4 (5 points)

In the latter part of the lab, we discussed calculating patch-based and landscape-based metrics of connectivity. Patch number 6 had the lowest degree (n = 2) while patch number 7 had one of the highest degrees (n = 12). First delete patch number 6 from the network, and then examine the impacts on the landscape level metrics integral index of connectivity (ICC) and probability of connectivity (PC). Now replace patch 6 back into the network and instead delete patch number 7. Again, examine how ICC and PC change. How does the loss of patch 6 compare to the loss of patch 7 in terms of its impacts on landscape-level connectivity?

```{r}
#whole
nodes = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/kite_nodes.csv')
area = nodes$area

coords = cbind(nodes$XCoord, nodes$YCoord)
distMat = pointDistance(coords, lonlat=F)
distMat = distMat / 1000 # in km

aObs = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/kite_movement.csv')[,-1]
diag(aObs) = 0

aMean = matrix(0, nrow=nrow(aObs), ncol=ncol(aObs))
aMean[distMat < aMeanDist] = 1
diag(aMean) = 0

aProb = matrix(0, nrow=nrow(aObs), ncol=ncol(aObs))
alpha = 1/aMeanDist
# Calculate weights in the distance matrix
aProb = exp(-alpha*distMat)
diag(aProb)=0

graphAmean = graph_from_adjacency_matrix(aMean, mode='undirected')
graphAprob = graph_from_adjacency_matrix(aProb, mode='undirected', weighted=T)
graphAobs = graph_from_adjacency_matrix(as.matrix(aObs), mode='directed', weighted=T)
AL = 63990 # Approximate study area in km^2

#Create a matrix of shortet paths
nlMat = distances(graphAmean)

#In cases where you have isolated patches, we assign an arbitrarily large value
nlMat [is.infinite(nlMat)] = 1000

iicMat = outer(area, area)/(1+nlMat)
iic = sum(iicMat)/AL^2
iic

pStarMat = distances(graphAprob, weights=E(graphAprob)$weight)

# Back-transform to probabilities
pStarMat = exp(-pStarMat)
# Numerator of PC
pcNum = outer(area, area)*pStarMat
# Probability of connectivity
pc = sum(pcNum)/AL^2
pc

### Minus Patch 6###
nodes2 = nodes[-6,]

coords = cbind(nodes2$XCoord, nodes2$YCoord)
distMat2 = pointDistance(coords, lonlat=F)
distMat2 = distMat / 1000 # in km

aObs2 = aObs[-6,-6]
distMat2 = distMat2[-6, -6]

aMean2 = matrix(0, nrow=nrow(aObs2), ncol=ncol(aObs2))
aMean2[distMat2 < aMeanDist] = 1
diag(aMean2) = 0

aProb2 = matrix(0, nrow=nrow(aObs2), ncol=ncol(aObs2))
alpha = 1/aMeanDist
# Calculate weights in the distance matrix
aProb2 = exp(-alpha*distMat2)
diag(aProb2)=0

graphAmean2 = graph_from_adjacency_matrix(aMean2, mode='undirected')
graphAprob2 = graph_from_adjacency_matrix(aProb2, mode='undirected', weighted=T)
graphAobs2 = graph_from_adjacency_matrix(as.matrix(aObs2), mode='directed', weighted=T)
AL = 63990 # Approximate study area in km^2

#Create a matrix of shortet paths
nlMat2 = distances(graphAmean2)

#In cases where you have isolated patches, we assign an arbitrarily large value
nlMat2 [is.infinite(nlMat2)] = 1000

area2 = area[-6]

iicMat2 = outer(area2, area2)/(1+nlMat2)
iic2 = sum(iicMat2)/AL^2
iic2

pStarMat2 = distances(graphAprob2, weights=E(graphAprob2)$weight)

# Back-transform to probabilities
pStarMat2 = exp(-pStarMat2)

# Numerator of PC
pcNum2 = outer(area2, area2)*pStarMat2

# Probability of connectivity
pc2 = sum(pcNum2)/AL^2
pc2

###Minus Patch 7###
nodes3 = nodes[-7,]

coords = cbind(nodes3$XCoord, nodes3$YCoord)
distMat3 = pointDistance(coords, lonlat=F)
distMat3 = distMat / 1000 # in km

aObs3 = aObs[-7,-7]
distMat3 = distMat3[-7,-7]

aMean3 = matrix(0, nrow=nrow(aObs3), ncol=ncol(aObs3))
aMean3[distMat3 < aMeanDist] = 1
diag(aMean3) = 0

aProb3 = matrix(0, nrow=nrow(aObs3), ncol=ncol(aObs3))
alpha = 1/aMeanDist
# Calculate weights in the distance matrix
aProb3 = exp(-alpha*distMat3)
diag(aProb3)=0

graphAmean3 = graph_from_adjacency_matrix(aMean3, mode='undirected')
graphAprob3 = graph_from_adjacency_matrix(aProb3, mode='undirected', weighted=T)
graphAobs3 = graph_from_adjacency_matrix(as.matrix(aObs3), mode='directed', weighted=T)
AL = 63990 # Approximate study area in km^2

#Create a matrix of shortet paths
nlMat3 = distances(graphAmean3)

#In cases where you have isolated patches, we assign an arbitrarily large value
nlMat3 [is.infinite(nlMat3)] = 1000

area3 = area[-7]

iicMat3 = outer(area3, area3)/(1+nlMat3)
iic3 = sum(iicMat3)/AL^2
iic3

pStarMat3 = distances(graphAprob3, weights=E(graphAprob3)$weight)

# Back-transform to probabilities
pStarMat3 = exp(-pStarMat3)

# Numerator of PC
pcNum3 = outer(area3, area3)*pStarMat3

# Probability of connectivity
pc3 = sum(pcNum3)/AL^2
pc3

```
*How does the loss of patch 6 compare to the loss of patch 7 in terms of its impacts on landscape-level connectivity?*

The integral index of connectivity (iic) looks at the amount of habitat in a space, with 0 = no habitat and 1 = all habitat. The iic for the original kite network is 0.0028, the -6 patch iic is 0.0035, and the -7 patch iic is 0.0033. I think the reason why the iic increased for both -6 patch and -7 patch is that datasets are smaller, and the patches must lower than average habitat which increase the iic for both -6 and -7 patches. I think the iic metric is not a decent connectivity metric for species found in very fragmented habitats. This could be due to habitat loss or preferences for naturally fragmented habitats like Florida which have many patches of many many different habitat types all next to each other. 

Probability of connectivity (pc) describes the probability that two random points in a landscape are reachable from one another given the habitat patch matrix and links between patches. The pc for the original kite network is 0.00595, the -6 patch pc is 0.00276, and the -7 patch pc is 0.00263. I think the decrease in pc between the original and the -6 and -7 patches are due to the lost connections and patches in these datasets. It is interesting how close -6 patch is to -7 patch given that -6 only had two connections and -7 has 12 connections. -6 patch must be important to connecting two larger networks given its larger than expected impacts. -7 also probably has redundant connections that make its presence less important than it first appears

$\color{red}{\text{Good. +5}}$