---
title: "R Notebook"
output: html_notebook
---

```{r, warning=F, message=F}

rm(list=ls())

require(sf)
require(tidyterra)
require(dismo)
require(tidyverse)
require(terra)
require(predicts)
require(ggnewscale)
require(mgcv)
require(randomForest)
require(maxnet)
require(enmSdmX)
require(gbm)
require(PresenceAbsence)
require(ecospat)
#Don't forget to load your other R packages!
```

# This first code chunk just recreates the maps we built in the lab.

```{r}

# Model building data
vathData = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week8/vath_2004.csv')

vathPres = vathData %>% filter(VATH==1)
vathAbs = vathData %>% filter(VATH==0)

vathPresXy = as.matrix(vathPres %>% dplyr::select(EASTING, NORTHING))
vathAbsXy = as.matrix(vathAbs %>% dplyr::select(EASTING, NORTHING))



# Validation data
vathVal = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week8/vath_VALIDATION.csv')

vathValPres = vathVal %>% filter(VATH==1)
vathValAbs = vathVal %>% filter(VATH==0)

vathValXy = as.matrix(vathVal %>% dplyr::select(EASTING, NORTHING))
vathValPresXy = as.matrix(vathValPres %>% dplyr::select(EASTING, NORTHING))
vathValAbsXy = as.matrix(vathValAbs %>% dplyr::select(EASTING, NORTHING))



# Bringing in the covariates
elev = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/elevation.tif')
canopy = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/canopy.tif')
mesic = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/mesic.tif')
precip = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/precip.tif')


# Resampling to make the covariate rasters match
mesic = resample(x = mesic, y = elev, 'near')
precip = resample(x = precip, y = elev, 'bilinear')

mesic = mask(mesic, elev)
precip = mask(precip, elev)

# Mesic forest within 1 km
probMatrix = focalMat(mesic, 1000, type='circle', fillNA=FALSE)
mesic1km = focal(mesic, probMatrix, fun='sum')


# Building the raster stack
layers = c(canopy, elev, mesic1km, precip)
names(layers) = c('canopy', 'elev', 'mesic1km', 'precip')


#Creating background points
set.seed(23)

backXy = data.frame(backgroundSample(layers, n=2000, p=vathPresXy))

# Extracting covariates for our different points
presCovs = extract(layers, vathPresXy)
absCovs = extract(layers, vathAbsXy)
backCovs = extract(layers, backXy)
valCovs = extract(layers, vathValXy)

presCovs = data.frame(vathPresXy, presCovs, pres=1)
absCovs = data.frame(vathAbsXy, absCovs, pres=0)
backCovs = data.frame(backXy, backCovs, pres=0)
valCovs = data.frame(vathValXy, valCovs)

presCovs = presCovs[complete.cases(presCovs),]
absCovs = absCovs[complete.cases(absCovs),]
backCovs = backCovs[complete.cases(backCovs),]

# Combining presence and background data into one dataframe

backCovs = backCovs %>%dplyr:: select(-ID)
colnames(presCovs)[1:2] = c('x', 'y')
colnames(absCovs)[1:2] = c('x', 'y')

presBackCovs = rbind(presCovs, backCovs)
presAbsCovs = rbind(presCovs, absCovs)

valCovs = valCovs %>% mutate(VATH = vathVal$VATH)
valCovs = valCovs[complete.cases(valCovs),]

```
```{r}
# Four models #

#1.  Fitting GLM
glmModel = glm(pres ~ canopy + elev + I(elev^2) + mesic1km + precip, family='binomial', data=presBackCovs)

glmMap = predict(layers, glmModel, type='response')


#2. Fitting GAM
gamModel = gam(pres ~ s(canopy, k=6) + s(elev, k=6) + s(mesic1km, k=6) + s(precip, k=6), family='binomial', data=presBackCovs, method='ML')

gamMap = predict(layers, gamModel, type='response')


#3. Fitting boosted regression tree model

boostModel = gbm(pres ~ elev + canopy + mesic1km + precip, distribution='bernoulli', n.trees=100, interaction.depth=2, shrinkage=0.1, bag.fraction=0.5, data=presBackCovs)

boostMap = predict(layers, boostModel, type='response')
boostMap = mask(boostMap, layers$canopy)


#4. Fitting random forest model

rfModel = randomForest(as.factor(pres) ~ canopy + elev + mesic1km + precip, data=presBackCovs, mtry=2, ntree=500, na.action = na.omit)

rfMap = predict(layers, rfModel, type='prob', index=2)

```



# Challenge 1 (4 points)

In the lab, we fit 6 SDMs. We then calculated discrimination statistics for all 6 and a calibration plot for 1 of them. Create calibration plots for the remaining 5 models, and then make a decision (based on your suite of discrimination statistics and calibration plots) about which of your SDMs is "best." Defend your answer.

```{r}
tmp = valCovs %>% mutate(VATH = vathVal$VATH)
tmp = tmp[complete.cases(tmp),]

valData = data.frame('ID' = 1:nrow(tmp)) %>% 
  mutate(obs = tmp$VATH,
         glmVal = predict(glmModel, tmp %>% select(canopy:precip), type='response'),
         gamVal = predict(gamModel, tmp %>% select(canopy:precip), type='response'),
         boostVal = predict(boostModel, tmp %>% select(canopy:precip), type='response'),
         rfVal = predict(rfModel, tmp %>% select(canopy:precip), type='prob')[,2],
         maxentVal = predict(maxentModel, tmp %>% select(canopy:precip), type='logistic')[,1])

#Comparison
summaryEval = data.frame(matrix(nrow=0, ncol=9))


nModels = ncol(valData)-2

#Individual model AUCs
#glm
for(i in 1:glmModel){
  auc1 = auc(glmVal, which.model = i)}

#gam
for(i in 1:gamModel){
  auc2 = auc(gamVal, which.model = i)}

#boost
for(i in 1:boostModel){
  auc3 = auc(boostVal, which.model = i)}

#rf
for(i in 1:rfModel){
  
  #AUC
  auc4 = auc(rfVal, which.model = i)}
  
#Combined data
for(i in 1:nModels){
  
  #AUC
  auc = auc(valData, which.model = i)
  
  #Find threshold to maximize Kappa
  kappaOpt = optimal.thresholds(valData, which.model = i, opt.methods=3)
  
  #Sensitivity
  sens = sensitivity(cmx(valData, which.model=i, threshold = kappaOpt[[2]]))
  
  #Specificity
  spec = specificity(cmx(valData, which.model = i, threshold = kappaOpt[[2]]))
  
  #True skill statistic
  tss = sens$sensitivity + spec$specificity - 1
  
  #Kappa
  kappa = Kappa(cmx(valData, which.model = i, threshold = kappaOpt[[2]]))
  
  #Correlation between predicted and realized values
  corr = cor.test(valData[,2], valData[,i+2])$estimate
  
  #Log likelihood
  ll = sum(log(valData[,i+2]*valData[,2] + (1-valData[,i+2]) * (1-valData[,2])))
  ll = ifelse(ll == '-Inf', sum(log(valData[,i+2] + 0.01)*valData[,2] + log((1-valData[,i+2]))*(1-valData[,2])), ll)
  
  #Put them all together and save the values
  summaryI = c(i, auc$AUC, corr, ll, kappaOpt[[2]], sens$sensitivity, spec$specificity, tss, kappa[[1]])
  summaryEval = rbind(summaryEval, summaryI)
}

summaryEval = summaryEval %>% 
  setNames(c('model', 'auc', 'corr', 'll', 'threshold', 'sens', 'spec', 'tss', 'kappa')) %>% 
  mutate(model = colnames(valData)[3:8])

summaryEval
#
#
#
#

#Calibrating GLM
calibration.plot(valData, which.model=2, N.bins=20, xlab='predicted', ylab='Observed', main='glm')

#Calibrating GAM
calibration.plot(valData, which.model=3, N.bins=20, xlab='predicted', ylab='Observed', main='gam')

#Calibrating boosted regression tree model
calibration.plot(valData, which.model=4, N.bins=20, xlab='predicted', ylab='Observed', main='gbm')

#Calibrating random forest model 
calibration.plot(valData, which.model=5, N.bins=20, xlab='predicted', ylab='Observed', main='randomForest')


```

*Which of these SDMs is the best? Defend your answer*

None of these models are "good" models because all the models are correctly predicting thrush location correctly at 50% or less which is, at best, has the same prediction ability as randomly selecting a point, and, at worst, worse at predicting presence than randomly selecting a point in the study site. The "best" model is the GLM model, but only because it has more than one point noticeably above the 0.5 threshold, which suggests slightly better predictive ability than random chance, and averages to around 0.5 overall. 

The GAM model is the only other model that averages around 0.5,with a couple points barely above 0.5, but it is not as good as the GLM because these points are barely above 0.5 and the largest reported predicted area, at 0.37, is only supported by 3 data points, and could be skewed given the predictive ability of the next highest point. 

No other models approach 0.5 in average predictive ability.

# Challenge 2 (4 points)

Each SDM we created uses a different algorithm with different assumptions. Because of this, ecologists frequently use "ensemble" approaches that aggregate predictions from multiple models in some way. Here we are going to create an ensemble model by calculating a weighted average of the predicted occupancy values at each pixel. We will calculate weights based on model AUC values to ensure that the models with the best AUC values have the most influence on the predicted values in the ensemble model.

Create a raster stack that combines the glmMap, gamMap, boostMap, and rfMap (hint use c()).

Next, create a vector of the AUC values for each model.

Lastly, use the weighted.mean() function in the terra package to create the new raster as a weighted average of the previous 4 rasters.

Plot the result, and explain why we left out the bioclim and Maxent models for this ensemble model.

```{r}
layers = c(glmMap,gamMap,boostMap,rfMap)
names(layers) = c('glmMap','gamMap','boostMap','rfMap')
rasterStack = cbind(layers)

AUCVect = c(auc1, auc2, auc3, auc4)
names(AUCVect)=c('glm','gam','boost','rf')
AUCStack = cbind(AUCVect)

ggplot(rasterStack)+
  aes(x = Model, y = Predict, group = layers) +
  geom_line(stat = "weighted_mean")

ggplot(AUCStack) + 
  aes(x = Model, y = Predict, group = AUCVect) +
  geom_line(stat = "weighted_mean")

```

We left the bioclim and maxent models out of the ensemble because these models are constructed differently from the 4 included models. Bioclim constructs a matrix from the data and covariates in order to predict presence in an area. The maxent model uses a logistic model to predict presence in an area. Both bioclim and maxent models result in a different style graphs for predictive models compared to regressions and bootstrap models (which assume linear correlation) and would reduce the predictive strength of the ensemble. 



# Challenge 3 (4 points)

Is this ensemble model an improvement over one of the models you built previously? Provide evidence and explain the criteria you used to come to your conclusion.

```{r}
set.seed(23)

nFolds = 5
kfoldPres = kfold(presCovs, k=nFolds)
kfoldBack = kfold(backCovs, k=nFolds)

boyceVals = rep(NA, nFolds)

for(i in 1:nFolds){
  valPres = presCovs[kfoldPres==i,]
  
  trainPres = presCovs[kfoldPres!=i,]
  trainBack = backCovs[kfoldBack!=i,]
  trainBoth = rbind(trainPres, trainBack)
  
  glmModel2 = glm(pres ~ canopy + elev + I(elev^2) + mesic1km + precip, family='binomial', data=trainBoth)

  valData = data.frame('ID' = 1:nrow(valPres)) %>% 
  mutate(obs = valPres$pres,
         glmVal = predict(glmModel2, valPres %>% select(canopy:precip), type='response'))
  
boyceVals[i] = ecospat.boyce(fit = glmMap, obs=valData[,3], res=100, PEplot=F)$cor

}

mean(boyceVals)

```

I was not able to compare the ensemble to the initial models, but considering the original models are bad models, and they all contained roughly similar predictive strength, I would not expect their ensemble to have better predictive strength. The predictive potential of an ensemble can only be as good as the models used to construct the prediction.



# Challenge 4 (4 points)

In the lab we built models using presence-background data then validated those models with presence-absence data. For this challenge, you're going to compare the predictive ability of a model built using presence-background data with one built using presence-absence data.

Fit a GLM using the presence-background data as we did in the lab (i.e., use the presBackCovs dataframe). Fit a second GLM using the presence-absence data (i.e., use the presAbsCovs dataframe). Validate both of these models on the novel presence-absence data (valCovs dataset). Specifically, calculate and compare AUC, Kappa, and TSS for these two models. Which model does a better job of prediction for the validation data and why do you think that is? 

```{r}
#presence-background data
set.seed(23)

backXy = data.frame(backgroundSample(layers, n=2000, p=vathPresXy))

presCovs = extract(layers, vathPresXy)
backCovs = extract(layers, backXy)
valCovs = extract(layers, vathValXy)

presCovs = data.frame(vathPresXy, presCovs, pres=1)
backCovs = data.frame(backXy, backCovs, pres=0)
valCovs = data.frame(vathValXy, valCovs)

presCovs = presCovs[complete.cases(presCovs),]
backCovs = backCovs[complete.cases(backCovs),]
valCovs = valCovs[complete.cases(valCovs),]


backCovs = backCovs %>% select(-ID)
colnames(presCovs)[1:2] = c('x', 'y')

presBackCovs = rbind(presCovs, backCovs)

#AUC, Kappa, TSS
summaryEval = data.frame(matrix(nrow=0, ncol=9))

nModels = ncol(valData)-2


for(i in 1:nModels){
  
  #AUC
  auc = auc(valData, which.model = i)
  
  #Find threshold to maximize Kappa
  kappaOpt = optimal.thresholds(valData, which.model = i, opt.methods=3)
  
  #Sensitivity
  sens = sensitivity(cmx(valData, which.model=i, threshold = kappaOpt[[2]]))
  
  #Specificity
  spec = specificity(cmx(valData, which.model = i, threshold = kappaOpt[[2]]))
  
  #True skill statistic
  tss = sens$sensitivity + spec$specificity - 1
  
  #Kappa
  kappa = Kappa(cmx(valData, which.model = i, threshold = kappaOpt[[2]]))
  
  #Correlation between predicted and realized values
  corr = cor.test(valData[,2], valData[,i+2])$estimate
  
  #Log likelihood
  ll = sum(log(valData[,i+2]*valData[,2] + (1-valData[,i+2]) * (1-valData[,2])))
  ll = ifelse(ll == '-Inf', sum(log(valData[,i+2] + 0.01)*valData[,2] + log((1-valData[,i+2]))*(1-valData[,2])), ll)
  
  #Put them all together and save the values
  summaryI = c(i, auc$AUC, corr, ll, kappaOpt[[2]], sens$sensitivity, spec$specificity, tss, kappa[[1]])
  summaryEval = rbind(summaryEval, summaryI)
}

summaryEval = summaryEval %>% 
  setNames(c('model', 'auc', 'corr', 'll', 'threshold', 'sens', 'spec', 'tss', 'kappa')) %>% 
  mutate(model = colnames(valData)[3:8])

summaryEval

#presence-absence data 
# Extracting covariates for our different points
presCovs = extract(layers, vathPresXy)
backCovs = extract(layers, backXy)
valCovs = extract(layers, vathValXy)

presCovs = data.frame(vathPresXy, presCovs, pres=1)
backCovs = data.frame(backXy, backCovs, pres=0)
valCovs = data.frame(vathValXy, valCovs)

presCovs = presCovs[complete.cases(presCovs),]
backCovs = backCovs[complete.cases(backCovs),]

# Combining presence and background data into one dataframe

backCovs = backCovs %>% dplyr:: select(-ID)
colnames(presCovs)[1:2] = c('x', 'y')

presBackCovs = rbind(presCovs, backCovs)

#AUC, Kappa, TSS
summaryEval = data.frame(matrix(nrow=0, ncol=9))

nModels = ncol(valData)-2


for(i in 1:nModels){
  
  #AUC
  auc = auc(valData, which.model = i)
  
  #Find threshold to maximize Kappa
  kappaOpt = optimal.thresholds(valData, which.model = i, opt.methods=3)
  
  #Sensitivity
  sens = sensitivity(cmx(valData, which.model=i, threshold = kappaOpt[[2]]))
  
  #Specificity
  spec = specificity(cmx(valData, which.model = i, threshold = kappaOpt[[2]]))
  
  #True skill statistic
  tss = sens$sensitivity + spec$specificity - 1
  
  #Kappa
  kappa = Kappa(cmx(valData, which.model = i, threshold = kappaOpt[[2]]))
  
  #Correlation between predicted and realized values
  corr = cor.test(valData[,2], valData[,i+2])$estimate
  
  #Log likelihood
  ll = sum(log(valData[,i+2]*valData[,2] + (1-valData[,i+2]) * (1-valData[,2])))
  ll = ifelse(ll == '-Inf', sum(log(valData[,i+2] + 0.01)*valData[,2] + log((1-valData[,i+2]))*(1-valData[,2])), ll)
  
  #Put them all together and save the values
  summaryI = c(i, auc$AUC, corr, ll, kappaOpt[[2]], sens$sensitivity, spec$specificity, tss, kappa[[1]])
  summaryEval = rbind(summaryEval, summaryI)
}

summaryEval = summaryEval %>% 
  setNames(c('model', 'auc', 'corr', 'll', 'threshold', 'sens', 'spec', 'tss', 'kappa')) %>% 
  mutate(model = colnames(valData)[3:8])

summaryEval

```

I think the presence-absence data would provide a better model for predicting the presence of thrushes because the absence data confirms locations where thrushes were not found. Background data estimates locations where birds were not found based on the locations where the birds were found. I think this misses more potential locations compared to the presence-absence data. However, I think neither model is necessarily good at predicting thrush presence.




# Do not work of Challenge 5 #
# Challenge 5 (4 points)

Now calculate the same statistics (AUC, Kappa, and TSS) for each model you developed in Challenge 4 using K-fold validation with 5 groups. Do these models perform better or worse based on K-fold validation (as compared to validation based on novel data)? Why might that occur?

```{r}
#Place your code here
```

Answer the question here.
