---
title: "R Notebook"
output: html_notebook
---
```{r}
require(tidyverse)
require(terra)
require(tidyterra)
require(sf)
require(adehabitatHR)
require(adehabitatLT)
```

```{r}


land = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week10/panther_landcover.tif')


classification = read.table('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week10/landcover%20reclass.txt', header=T) 

head(classification)

unique(classification$Description2)

land = classify(land, classification[,c(1,3)])
land = categories(land, value=unique(classification[,c(3,4)]))
plot(land)

panthers = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week10/panthers.shp') %>% 
  mutate(CatID = as.factor(CatID))

summary(panthers)
unique(panthers$CatID)

```

```{r}
#Map
crs(land, proj=T) == crs(panthers, proj=T)


ggplot()+
  geom_spatraster(data = land, aes(fill=Description2))+
  scale_fill_manual(values = terrain.colors(14))+
  geom_sf(data = panthers, aes(shape=CatID))

#Wet forest
wetForest = land
values(wetForest) = 0
wetForest[land %in% c(10,12)] = 1

probMatrix = focalMat(wetForest, 5000, type='circle', fillNA=FALSE)
wetFocal = focal(wetForest, probMatrix, fun='sum', na.rm=T)


#Dry forest
dryForest = land
values(dryForest) = 0
dryForest[land %in% c(11, 13)] = 1

probMatrix = focalMat(dryForest, 5000, type='circle', fillNA=FALSE)
dryFocal = focal(dryForest, probMatrix, fun='sum', na.rm=T)

layers = c(land, wetFocal, dryFocal)
names(layers) = c('landcover', 'wetForest', 'dryForest')
plot(layers)
```



# Challenge 1 (4 points)

Create a table with 6 rows and 5 columns. Each row should represent one panther. The first column should represent the panther's ID, and the other 4 columns should represent the estimated 95% home range from the 4 methods we used (MCP, KDE, a-LoCoH, and Brownian bridge). Make sure that all of your units for your home range estimates are the same. Ensure that the final table outputs from the code below.

```{r}
panthersSp = as(panthers, 'Spatial')

#Panthers 

panther100 = panthersSp[panthersSp$CatID==100,]
panther130 = panthersSp[panthersSp$CatID==130,]
panther131 = panthersSp[panthersSp$CatID==131,]
panther137 = panthersSp[panthersSp$CatID==137,]
panther143 = panthersSp[panthersSp$CatID==143,]
panther147 = panthersSp[panthersSp$CatID==147,]

#95% MCP
mcp95 = mcp(panthersSp[,'CatID'], percent = 95, unin='m', unout='km2')
mcp95Sf = st_as_sf(mcp95)
mcp95

#95% KDE
kernelHrefEpa = kernelUD(panthersSp[,'CatID'], h='href', kern='epa')
kernelHrefEpa[[2]]@h
kernel.area(kernelHrefEpa)
kde95Sf = st_as_sf(getverticeshr(kernelHrefEpa, percent=95))
```

```{r}
#95% a-LoCoH
aInit100 = round(max(dist(coordinates(panther100))),0)
aInit130 = round(max(dist(coordinates(panther130))),0)
aInit131 = round(max(dist(coordinates(panther131))),0)
aInit137 = round(max(dist(coordinates(panther137))),0)
aInit143 = round(max(dist(coordinates(panther143))),0)
aInit147 = round(max(dist(coordinates(panther147))),0)

aSearch100 = seq(aInit100, 2*aInit100, by=3000) #distance in m
aSearch130 = seq(aInit130, 2*aInit130, by=3000) #distance in m
aSearch131 = seq(aInit131, 2*aInit131, by=3000) #distance in m
aSearch137 = seq(aInit137, 2*aInit137, by=3000) #distance in m
aSearch143 = seq(aInit143, 2*aInit143, by=3000) #distance in m
aSearch147 = seq(aInit147, 2*aInit147, by=3000) #distance in m

locohArange100 = LoCoH.a.area(SpatialPoints(coordinates(panther100)), arange=aSearch)
locohArange130 = LoCoH.a.area(SpatialPoints(coordinates(panther130)), arange=aSearch)
locohArange131 = LoCoH.a.area(SpatialPoints(coordinates(panther131)), arange=aSearch)
locohArange137 = LoCoH.a.area(SpatialPoints(coordinates(panther137)), arange=aSearch)
locohArange143 = LoCoH.a.area(SpatialPoints(coordinates(panther143)), arange=aSearch)
locohArange147 = LoCoH.a.area(SpatialPoints(coordinates(panther147)), arange=aSearch)
```

```{r}

aSearch100[5]
aSearch130[5]
aSearch131[5]
aSearch137[5]
aSearch143[5]
aSearch147[5]

locohA100 = LoCoH.a(SpatialPoints(coordinates(panther100)), a=aSearch[5])
locohA130 = LoCoH.a(SpatialPoints(coordinates(panther130)), a = aSearch[5])
locohA131 = LoCoH.a(SpatialPoints(coordinates(panther131)), a=aSearch[5])
locohA137 = LoCoH.a(SpatialPoints(coordinates(panther137)), a = aSearch[5])
locohA143 = LoCoH.a(SpatialPoints(coordinates(panther143)), a=aSearch[5])
locohA147 = LoCoH.a(SpatialPoints(coordinates(panther147)), a = aSearch[5])

locohA100Sf = st_as_sf(locohA100)
locohA130Sf = st_as_sf(locohA130)
locohA131Sf = st_as_sf(locohA131)
locohA137Sf = st_as_sf(locohA137)
locohA143Sf = st_as_sf(locohA143)
locohA147Sf = st_as_sf(locohA147)

locohA92Sf100 = locohA100Sf[97,]
locohA92Sf130 = locohA130Sf[97,]
locohA92Sf131 = locohA131Sf[97,]
locohA92Sf137 = locohA137Sf[97,]
locohA92Sf143 = locohA143Sf[97,]
locohA92Sf147 = locohA147Sf[97,]

locohA92Sf = c(locohA92Sf100,locohA92Sf130,locohA92Sf131,locohA92Sf137,locohA92Sf143,locohA92Sf147)
```

```{r}
#95% Brownian bridge
substrRight = function(x, n){
  substr(x, nchar(x) - n+1, nchar(x))
}

panthersSp = panthers %>% 
  mutate(Juldate = as.character(Juldate)) %>% 
  mutate(date = as.numeric(substrRight(Juldate, 3))) %>% 
  mutate(Date = as.Date(date, origin=as.Date("2006-01-01"))) %>% 
  mutate(Date = as.POSIXct(Date, "%Y-%m-%d")) %>% 
  as('Spatial')

pantherLtraj = as.ltraj(xy=coordinates(panthersSp), date=panthersSp$Date, id=panthersSp$CatID, typeII=T)
plot(pantherLtraj)

sigma1 = liker(pantherLtraj, sig2=450, rangesig1=c(2, 100))
sigma1
```

```{r}
bb100 = kernelbb(pantherLtraj[1], sig = 7.2, sig2 = 450, grid = 500)
bb130 = kernelbb(pantherLtraj[2], sig = 7.2, sig2 = 450, grid = 500)
bb131 = kernelbb(pantherLtraj[3], sig = 7.2, sig2 = 450, grid = 500)
bb137 = kernelbb(pantherLtraj[4], sig = 7.2, sig2 = 450, grid = 500)
bb143 = kernelbb(pantherLtraj[5], sig = 7.2, sig2 = 450, grid = 500)
bb147 = kernelbb(pantherLtraj[6], sig=7.2, sig2=450, grid=500)

bb100sf = st_as_sf(getverticeshr(bb100, percent = 95))
bb130sf = st_as_sf(getverticeshr(bb130, percent = 95))
bb131sf = st_as_sf(getverticeshr(bb131, percent = 95))
bb137sf = st_as_sf(getverticeshr(bb137, percent = 95))
bb143sf = st_as_sf(getverticeshr(bb143, percent = 95))
bb147sf = st_as_sf(getverticeshr(bb147, percent=95))

IDs = c(panther100, panther130, panther131, panther137, panther143, panther147)
Brownian = c(bb100sf, bb130sf, bb131sf, bb137sf, bb143sf, bb147sf)
Table1 = cbind(IDs, Brownian,locohA92Sf, kde95Sf, mcp95Sf)  

```


No text necessary.


# Challenge 2 (4 points)

Build a boxplot showing the distribution of home range sizes. Your x-axis should have 4 categories (one for each home range type), and your y-axis should show the distribution of the home ranges for the 6 panthers. Which home range estimates tend to be largest? Which ones tend to be smallest? What is your intuitive explanation for why this pattern occurs?

```{r}
newdata = pivot_longer(dataset,cols = Brownian,locohA92Sf, kde95Sf, mcp95Sf)
ggplot(newdata,aes(x = Brownian,locohA92Sf,kde95Sf, mcp95Sf, y = IDs)) + 
  geom_boxplot()

```

The home ranges tend to be the largest for Panther 130 and Panther 147, which is likely due to these ranges containing predominately wetland or upland habitats respectively. Panther 100 and Panther 143 have the smallest home ranges. These panthers live on edge areas between areas dominated by wetlands and areas dominated by uplands. I think that Florida Panthers prefer habitats that include large patches of wetlands and large patches of uplands. 

# Challenge 3 (3 points)

Choose 1 panther and generate a 100% home range based on the MCP. Calculate (and report) the proportion of this polygon comprised of wet and dry forest. Do you think these amounts are representative of the wet and dry forest available to your panther? Why or why not?

```{r}
panthersSp = as(panthers, 'Spatial')
mcp130 = mcp(panther130, percent = 100, unin='m', unout='km2')
mcp130Sf = st_as_sf(mcp130)
mcp130

ggplot()+
  geom_spatraster(data = land, aes(fill=Description2))+
  scale_fill_manual(values = terrain.colors(14))+
  geom_sf(data = panthers, aes(shape=CatID==130))+
  geom_sf(data = mcp130Sf, alpha=0.5)+
  ggtitle('100% MCP')

Cat130wet = extract(wetForest, mcp130Sf)
410/6695
Cat130dry = extract(dryForest, mcp130Sf)
1040/6695
sum(Cat130wet$Description2)
sum(Cat130dry$Description2)
Cat130wet
Cat130dry

```
Panther 130 was found in wet forests 6% of the time and found in dry forests 15% of the time. Considering the higher amount of wet forest habitat in 130's range, this panther is found is wet forest far less than random distribution, and is found in dry forest more than random distribution.

# Challenge 4 (6 points)

Using the same panther you chose for Challenge 3, generate 10 sf objects, one representing each of the 10%, 20%, 30%, ..., 90%, and 100% home ranges based on a Brownian bridge model. Extract the proportion of each of those territories comprised of dry forest and wet forest. Now generate two line graphs, one for wet forest and one for dry. On the x-axis, plot increasing home range size (10-100%). On the y-axis plot the proportion of the territory comprised of wet/dry forest, and link these with a line. Finally, add a horizontal line that shows the rough proportion of that forest type available to the panther (based on challenge 3). Show your output below.

```{r}
substrRight = function(x, n){
  substr(x, nchar(x) - n+1, nchar(x))
}

panthersSp = panthers %>% 
  mutate(Juldate = as.character(Juldate)) %>% 
  mutate(date = as.numeric(substrRight(Juldate, 3))) %>% 
  mutate(Date = as.Date(date, origin=as.Date("2006-01-01"))) %>% 
  mutate(Date = as.POSIXct(Date, "%Y-%m-%d")) %>% 
  as('Spatial')

pantherLtraj = as.ltraj(xy=coordinates(panthersSp), date=panthersSp$Date, id=panthersSp$CatID, typeII=T)

plot(pantherLtraj)

sigma1 = liker(pantherLtraj, sig2=450, rangesig1=c(2, 100))
sigma1

bb130 = kernelbb(pantherLtraj[2], sig=7.2, sig2=450, grid=500)
plot(bb130)

bb10Sf = st_as_sf(getverticeshr(bb130, percent=10))
bb20Sf = st_as_sf(getverticeshr(bb130, percent=20))
bb30Sf = st_as_sf(getverticeshr(bb130, percent=30))
bb40Sf = st_as_sf(getverticeshr(bb130, percent=40))
bb50Sf = st_as_sf(getverticeshr(bb130, percent=50))
bb60Sf = st_as_sf(getverticeshr(bb130, percent=60))
bb70Sf = st_as_sf(getverticeshr(bb130, percent=70))
bb80Sf = st_as_sf(getverticeshr(bb130, percent=80))
bb90Sf = st_as_sf(getverticeshr(bb130, percent=90))
bb100Sf = st_as_sf(getverticeshr(bb130, percent=100))



```

# Challenge 5 (3 points)

Does the panther you chose in Challenge 3/4 prefer dry and/or wet forest? What is your justification for your answer?

Panther 130 prefers drier forests, despite being in a sub-optimal location. It spends most of its time and movement in the drier sections of its home range, which are hardwood swamps and upland forests. Florida is a very low-lying state, and Florida panther habitat has few dry or upland forests. hardwood swamp is second driest habitat these cats have access to because these patches are raised up by the roots of the hardwoods growing on these patches, despite technically being considered wetland patches. The upper part of 130's home range contains more water and wetter wetland types compared to the lower part of 130's range.  

