---
title: "Lab 8"
output: html_document
date: "2024-03-11"
---
```{r, warning=F, message=F}
require(tidyterra)
require(dismo)
require(tidyverse)
require(terra)
require(predicts)
require(ggnewscale)
require(mgcv)
require(randomForest)
require(maxnet)
require(enmSdmX)
require(gbm)
require(landscapemetrics)
```

```{r}
#Importing bird data#


vathData = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week8/vath_2004.csv')

vathPres = vathData %>% filter(VATH==1)
vathAbs = vathData %>% filter(VATH==0)

vathPresXy = as.matrix(vathPres %>% select(EASTING, NORTHING))
vathAbsXy = as.matrix(vathAbs %>% select(EASTING, NORTHING))
```

```{r}

vathVal = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week8/vath_VALIDATION.csv')

vathValPres = vathVal %>% filter(VATH==1)
vathValAbs = vathVal %>% filter(VATH==0)

vathValXy = as.matrix(vathVal %>% select(EASTING, NORTHING))
vathValPresXy = as.matrix(vathValPres %>% select(EASTING, NORTHING))
vathValAbsXy = as.matrix(vathValAbs %>% select(EASTING, NORTHING))

```

```{r}
#Prepping data for Part 1#
elev = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/elevation.tif')
canopy = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/canopy.tif')
mesic = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/mesic.tif')
precip = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/precip.tif')

crs(elev) = crs(mesic)
crs(canopy) = crs(mesic)
```

```{r}
compareGeom(elev, canopy, stopOnError=F)
compareGeom(elev, precip, stopOnError=F)
compareGeom(elev, mesic, stopOnError=F)
```

```{r}
elev

mesic

ggplot()+
  geom_raster(data=elev, aes(x=x, y=y, fill=elev_km))+
  scale_fill_gradientn(colors=c('blue', 'white', 'red'))+
  new_scale_fill()+
  geom_raster(data=mesic, aes(x=x, y=y, fill=a_pmesic), alpha=0.2)+
  scale_fill_gradientn(colors=c('yellow', 'black', 'green'))+
  coord_fixed()+
  theme_bw()+
  theme(panel.grid=element_blank())+
  xlim(21000, 23000)+
  ylim(302000, 303000)

```

```{r}
mesic = resample(x = mesic, y = elev, 'near')
precip = resample(x = precip, y = elev, 'bilinear')

mesic = mask(mesic, elev)
precip = mask(precip, elev)
```

```{r}
compareGeom(elev, precip, canopy, mesic)
```

# Challenge 1 (4 points)

In the lab, we created 6 species distribution models (SDMs) for the same species using 6 different techniques. Plot the maps generated from (1) the bioclim envelope function, (2) the GLM model, and (3) the random forest model next to one another.

```{r cars}
probMatrix = focalMat(mesic, 1000, type='circle', fillNA=FALSE)
mesic1km = focal(mesic, probMatrix, fun='sum')

layers = c(canopy, elev, mesic, mesic1km, precip)
names(layers) = c('canopy', 'elev', 'mesic', 'mesic1km', 'precip')
plot(layers)
```

```{r}
pairs(layers, maxpixels=1000)

layers = c(canopy, elev, mesic1km, precip)
names(layers) = c('canopy', 'elev', 'mesic1km', 'precip')
```
```{r}
set.seed(23)

backXy = data.frame(backgroundSample(layers, n=2000, p=vathPresXy))

ggplot()+
  geom_raster(data=elev, aes(x=x, y=y, fill=elev_km))+
  geom_point(data=backXy, aes(x=x, y=y))+
  geom_point(data=vathPres, aes(x=EASTING, y=NORTHING), color='red', alpha=0.3)+
  coord_fixed()
```

```{r}
presCovs = extract(layers, vathPresXy)
backCovs = extract(layers, backXy)
valCovs = extract(layers, vathValXy)

presCovs = data.frame(vathPresXy, presCovs, pres=1)
backCovs = data.frame(backXy, backCovs, pres=0)
valCovs = data.frame(vathValXy, valCovs)

presCovs = presCovs[complete.cases(presCovs),]
backCovs = backCovs[complete.cases(backCovs),]
valCovs = valCovs[complete.cases(valCovs),]


backCovs = backCovs %>% select(-ID)
colnames(presCovs)[1:2] = c('x', 'y')

presBackCovs = rbind(presCovs, backCovs)

```

```{r}
#Envelope
tmp = presCovs %>% select(elev, precip, mesic1km, canopy) %>% 
  as.matrix()

bioclim = envelope(tmp)


plot(bioclim, a=1, b=2, p=0.95)
plot(bioclim, a=1, b=3, p=0.95)
plot(bioclim, a=3, b=4, p=0.95)

bioclimMap = predict(layers, bioclim)
plot(bioclimMap)

#GLM
glmModel = glm(pres ~ canopy + elev + I(elev^2) + mesic1km + precip, family='binomial', data=presBackCovs)

summary(glmModel)

glmMap = predict(layers, glmModel, type='response')
plot(glmMap)

#Random forest model
tuneRF(y = as.factor(presBackCovs$pres), x=presBackCovs[,3:6], stepFactor = 2, ntreeTry = 500)

rfModel = randomForest(as.factor(pres) ~ canopy + elev + mesic1km + precip, data=presBackCovs, mtry=2, ntree=500, na.action = na.omit)

rfMap = predict(layers, rfModel, type='prob', index=2)
plot(rfMap)
```
*What similarities and differences do you notice among these maps? What might explain some of these differences?*

Both the Envelope and Random forest model use the same bird presence scale, which maxes out a little above 80% presence (like 81-82). The GLM model has a scale that maxes at 37% presence, so this plot contains more higher presence gradients. All of the models suggest higher thrush presence is highly correlated to higher percent canopy and a particular elevation pattern, where thrush are found around slopes adjacent to lower elevation. 

The Envelope model excludes thrush presence from more high elevation areas, which is likely due to how Envelope models calculates presence through percentiles from observed environmental covariates where a thrush was detected. This also explains why model suggests higher thrush presence than the other models. Most of the areas have some similar observed environmental covariates compared to areas of known thrush presence. 

The GLM model is a much more conservative model compared to the Envelope and Random forest model. This could be due to the GLM model's limitation of assuming linear and/or polynomial relationships between bird presence and the covariates of interest. Based on the table provided for this GLM, elevation, precipation, and canopy are the most statistically significant covariates in to determine thrush presence. However, precipation and canopy probably have an interaction, so they could be colinear. 

The Random forest model uses classification of trees in order to build the model via the number of explanatory variables sampled for each tree and the estimated number of trees in the forest. Similar to the GLM model, the Random forest model that is plotted above is likely a conservative model in terms of thrush presence. This is due to the number of explanatory variables and number of trees used in the model, which are 2 and 500 respectively. However, changing these variables could easily far overestimate or underestimate thrush presence. Having an idea of reasonable thrush presence and how they interact with the explanatory variables is important for the bootstrap models like the Random forest model. 


# Challenge 2 (4 points)

When we fit our GLM in lab, we used background points, rather than true absence points, to represent pseudo-absences. Fit the exact same GLM model, only this time use presence and true absence data. That is, replace the background rows in the dataframe with rows that represent actual sites where surveys were completed but Varied Thrush were not detected. Once you've fit the GLM, build a new SDM from this fitted model and visually.
```{r}
#presence-background GLM
glmModel = glm(pres ~ canopy + elev + I(elev^2) + mesic1km + precip, family='binomial', data=presBackCovs)

summary(glmModel)

glmMap = predict(layers, glmModel, type='response')
plot(glmMap)

#setting-up absence
set.seed(23)

backXy1 = data.frame(vathValAbsXy)

ggplot()+
  geom_raster(data=elev, aes(x=x, y=y, fill=elev_km))+
  geom_point(data=backXy1, aes(x=EASTING, y=NORTHING))+
  geom_point(data=vathPres, aes(x=EASTING, y=NORTHING), color='red', alpha=0.3)+
  coord_fixed()
presCovs1 = extract(layers, vathPresXy)
backCovs1 = extract(layers, backXy1)
valCovs1 = extract(layers, vathValXy)

presCovs1 = data.frame(vathPresXy, presCovs1, pres=1)
backCovs1 = data.frame(backXy1, backCovs1, pres=0)
valCovs1 = data.frame(vathValXy, valCovs1)

presCovs1 = presCovs1[complete.cases(presCovs1),]
backCovs1 = backCovs1[complete.cases(backCovs1),]
valCovs1 = valCovs1[complete.cases(valCovs1),]

backCovs1 = backCovs1 %>% select(-ID)
colnames(presCovs1)[1:2] = c('EASTING', 'NORTHING')

presBackCovs1 = rbind(presCovs1, backCovs1)

#presence-absence GLM
glmModel1 = glm(pres ~ canopy + elev + I(elev^2) + mesic1km + precip, family='binomial', data=presBackCovs1)

summary(glmModel1)

glmMap1 = predict(layers, glmModel1, type='response')
plot(glmMap1)
```
*Compare the prediction surface to that built based on the presence-background model. What discrepancies do you notice, and what is your intuition regarding which of these two models is more reliable?*

The plot of the GLM presence-absence model has a larger scale of thrush presence compared to the GLM presence-background model. Additionally, the presence-absence model did map has more blurred looking presence coloring compared to the presence-background model, which suggests more of a presence gradient in the presence-absence model. 

I don't really know much about bird distributions, but based on my intuition I think that presence-absence model is more reliable for determining thrush presence. species are more likely to be found in optimal habitat, but they are likely to have infrequent presence all over the map since it is close to optimal habitat. 


# Challenge 3 (4 points)

Now plot the relationship between the 4 explanatory variables and the predicted occupancy values based on the two fitted GLM models (presence-background and presence-absence). Recall that we did this in the latter part of our lab. 


```{r}
#presence-background 
tmp = expand.grid(elev = seq(min(backCovs$elev), max(backCovs$elev), length=1000),
                  canopy = mean(backCovs$canopy),
                  precip = mean(backCovs$precip),
                  mesic1km = mean(backCovs$mesic1km))

elevData = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  cbind(tmp) %>% 
  select(bioclim:elev) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'elevation')

tmp = expand.grid(elev = mean(backCovs$elev),
                  canopy = seq(min(backCovs$canopy), max(backCovs$elev), length=1000),
                  precip = mean(backCovs$precip),
                  mesic1km = mean(backCovs$mesic1km))

canopyData = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  cbind(tmp) %>% 
  select(bioclim:maxent, canopy) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'canopy')

tmp = expand.grid(elev = mean(backCovs$elev),
                  canopy = mean(backCovs$canopy),
                  precip = seq(min(backCovs$precip), max(backCovs$precip), length=1000),
                  mesic1km = mean(backCovs$mesic1km))

precipData = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  cbind(tmp) %>% 
  select(bioclim:maxent, precip) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'precipitation')

tmp = expand.grid(elev = mean(backCovs$elev),
                  canopy = mean(backCovs$canopy),
                  precip = mean(backCovs$precip),
                  mesic1km = seq(min(backCovs$mesic1km), max(backCovs$mesic1km), length=1000))

mesicData = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  cbind(tmp) %>% 
  select(bioclim:maxent, mesic1km) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'mesic1km')


colnames(elevData)[1] = colnames(canopyData)[1] = colnames(precipData)[1] = colnames(mesicData)[1] = 'xValue'

tmp = rbind(elevData, canopyData, precipData, mesicData)

ggplot(tmp, aes(x=xValue, y=value, color=name))+
  facet_wrap(~variable, scales='free_x')+
  geom_line()+
  theme_bw()+
  theme(panel.grid=element_blank())

#presence-absence
tmp = expand.grid(elev = seq(min(backCovs1$elev), max(backCovs1$elev), length=1000),
                  canopy = mean(backCovs1$canopy),
                  precip = mean(backCovs1$precip),
                  mesic1km = mean(backCovs1$mesic1km))

elevData1 = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  
  cbind(tmp) %>% 
  select(bioclim:elev) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'elevation')

tmp = expand.grid(elev = mean(backCovs1$elev),
                  canopy = seq(min(backCovs1$canopy), max(backCovs1$elev), length=1000),
                  precip = mean(backCovs1$precip),
                  mesic1km = mean(backCovs1$mesic1km))

canopyData1 = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  cbind(tmp) %>% 
  select(bioclim:maxent, canopy) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'canopy')

tmp = expand.grid(elev = mean(backCovs1$elev),
                  canopy = mean(backCovs1$canopy),
                  precip = seq(min(backCovs1$precip), max(backCovs1$precip), length=1000),
                  mesic1km = mean(backCovs1$mesic1km))

precipData1 = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  cbind(tmp) %>% 
  select(bioclim:maxent, precip) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'precipitation')

tmp = expand.grid(elev = mean(backCovs1$elev),
                  canopy = mean(backCovs1$canopy),
                  precip = mean(backCovs1$precip),
                  mesic1km = seq(min(backCovs1$mesic1km), max(backCovs1$mesic1km), length=1000))

mesicData1 = data.frame(bioclim = predict(bioclim, tmp),
                 glm = predict(glmModel, tmp, type='response'),
                 gam = predict(gamModel, tmp, type='response'),
                 boost = predict(boostModel, tmp, type='response'),
                 rf = predict(rfModel, tmp, type='prob')[,2],
                 maxent = predict(maxentModel, tmp, type='logistic')[,1]) %>% 
  cbind(tmp) %>% 
  select(bioclim:maxent, mesic1km) %>% 
  pivot_longer(bioclim:maxent) %>% 
  mutate(variable = 'mesic1km')


colnames(elevData1)[1] = colnames(canopyData1)[1] = colnames(precipData1)[1] = colnames(mesicData1)[1] = 'xValue'

tmp = rbind(elevData1, canopyData1, precipData1, mesicData1)

ggplot(tmp, aes(x=xValue, y=value, color=name))+
  facet_wrap(~variable, scales='free_x')+
  geom_line()+
  theme_bw()+
  theme(panel.grid=element_blank())


```
*Do you notice any differences in the covariate patterns between the two models? Does this help you interpret the discrepancies between the predicted surfaces from the two models?*

The presence-absence model has a lower thrush presence scale. This makes the presence thresholds look larger for all the models. Additionally, some models such as the boost and rf models additional disprepancies for the covariates. Comparing the two models suggests that presence-absence data provides more detail on the presence data due to the lower scales. presence-background data could be more useful than the presence-absence model, but it depends on what aspects of thrush presence we are studying. 


# Challenge 4 (4 points)

Varied Thrush are considered forest-dependent, and thus one might characterize mesic forests as "habitat" for the species. Calculate the total amount of mesic forest in the study area, and the mean size of the mesic forest patches.

Using the SDM built from the random forest model, *convert the landscape into "habitat" and "non-habitat."* To do this, choose a threshold value in your SDM and convert all cells with predicted outcomes greater than this threshold to 1 and all cells with predicted values below your threshold to 0. Justify your choice of your threshold value. Now calculate the total amount of habitat and mean size of habitat patches based on this new raster (i.e., create patches of "habitat" based on aggregations of cells you deemed 1). 

```{r}
mesic1 = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week8/mesic.tif')

mesic1 = resample(x = mesic1, y = elev, 'near')

mesic1 = mask(mesic1, elev)
              
plot(mesic1)
lsm_l_ta(mesic1,directions = 8)
lsm_c_area_mn(mesic1,directions = 8)
lsm_c_np(mesic1,directions = 8)

```
# Random forest model

```{r}
tuneRF(y = as.factor(presBackCovs$pres), x=presBackCovs[,3:6], stepFactor = 2, ntreeTry = 500)

rfModel = randomForest(as.factor(pres) ~ canopy + elev + mesic1km + precip, data=presBackCovs, mtry=2, ntree=500, na.action = na.omit)

rfModelSimple = rfModel
rfModelSimple[rfModelSimple %in% c(0:0.2)] = 0 #
rfModelSimple[rfModelSimple %in% c(0.21:0.8)] = 1 #


#Give these numbers category names
tmp = data.frame('ID' = c(0,1),
                 'category' = c('absent', 'presence'))
#rfModelSimple = categories(rfModelSimple, value=tmp)
rfMap = predict(layers, rfModel, type='prob', index=2)

#rfMap = predict(layers, rfModelSimple, type='prob', index=2)
plot(rfMap)

#lsm_l_ta(rfModel,directions = 8)
#lsm_c_area_mn(rfModel,directions = 8)
#lsm_c_np(rfModel,directions = 8)
```
*How do the habitat amount and patch size values compare between the mesic forest approach and the SDM-based approach? In what situations might you rely on one map over the other?*

The SDM-based approach should have more total area than the mesic forest approach because there are birds seen outside of the mesic forests. However, the patch areas are likely smaller based on the appearance of both maps. Each model can provide answer to where to find a thrush, but they answer different aspects of that question. If you are just looking for a thrush, then the SDM works better, since the mesic forest misses sites that may have a thrush. If you are looking for presence density (where do these birds congregate) or individual presence over time, then mesic forest is likely more useful than SDM because they spend more time in mesic forests. Birds outside of mesic forests are likely in transit for resources or behaviors. 

# Challenge 5 (4 points)

When we fit the Maxent model in the lab, we used a regularization constant of 1. Fit the model two more times, using regularization (regmult) constants of 0.5 and 3. Construct figures showing the relationship between the 4 explanatory variables and the predicted outcome from these 3 fitted Maxent models. 

# Maxent models
```{r}
#regmult = 0.5
pbVect = presBackCovs$pres
covs = presBackCovs %>% select(canopy:precip)

maxentModel = maxnet(p = pbVect,
                     data= covs,
                     regmult = 0.5,
                     classes='lqpht')

plot(maxentModel, type='logistic')

maxentMap = predictMaxNet(maxentModel, layers, type='logistic')

par(mfrow=c(1,1))
plot(maxentMap)

#regmult = 1
pbVect = presBackCovs$pres
covs = presBackCovs %>% select(canopy:precip)

maxentModel = maxnet(p = pbVect,
                     data= covs,
                     regmult = 1,
                     classes='lqpht')

plot(maxentModel, type='logistic')

#regmult = 3
maxentMap = predictMaxNet(maxentModel, layers, type='logistic')

par(mfrow=c(1,1))
plot(maxentMap)

pbVect = presBackCovs$pres
covs = presBackCovs %>% select(canopy:precip)

maxentModel = maxnet(p = pbVect,
                     data= covs,
                     regmult = 3,
                     classes='lqpht')

plot(maxentModel, type='logistic')

maxentMap = predictMaxNet(maxentModel, layers, type='logistic')

par(mfrow=c(1,1))
plot(maxentMap)
```
*What is the regularization constant doing? Hint: you may need to Google it.*
regmult refers to the regularization multiplier that add new constraints to the model, which shrinks the betas to values that fit a balance between fit and complexity. This allows the maxent model to balance for both precision and generality when modeling. 

