---
title: "Lab_4_Landcover"
output: html_document
date: "2024-02-05"
---
```{r, warning=F, error=F, message=F}
rm(list=ls())
require(Voss)
require(tidyverse)
require(terra)
require(FedData)
require(sf)
require(tidyterra)
require(landscapemetrics)
```

```{r}
studyArea = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week4/studyArea.shp')
nlcd = get_nlcd(studyArea, label='AlLandscape', year=2011)

# values(nlcd) = as.character(values(nlcd))

plot(nlcd)
```

```{r}

nlcdSimple = nlcd
nlcdSimple[nlcdSimple==11] = 1 #Wet areas are a 1 now
nlcdSimple[nlcdSimple %in% c(21, 22, 23, 24)] = 2 #All developed areas are 2
nlcdSimple[nlcdSimple %in% c(31, 52)] = 3 #Barren land and shrub/scrub are 3
nlcdSimple[nlcdSimple %in% c(41,42,43)] = 4 #All forest types are 4
nlcdSimple[nlcdSimple == 71] = 5 #Grassland is 5
nlcdSimple[nlcdSimple %in% c(81,82)] = 6 #And agriculture is 6

#Give these numbers category names
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcdSimple = categories(nlcdSimple, value=tmp)

#And plot the new raster
ggplot(nlcdSimple, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))
```

```{r}
res(nlcdSimple)
ext(nlcdSimple)
levels(nlcdSimple)
```

```{r}

forest = nlcdSimple %>% 
  setValues(NA)
  
forest[nlcdSimple ==4] = 1

plot(forest)

```

```{r}
forestPatchId = patches(forest, directions=8, zeroAsNA=T, allowGaps=F)

plot(forestPatchId)
```


## Challenge 1 (4 points)

**The landscapemetrics package has functions for calculating 12 patch-level metrics. Calculate all 12 of these for every forest patch in our nlcdSimple raster (using the 8-direction rule). This will result in 12 unique values associated with each forest patch. Use the chart.Correlation() function from the PerformanceAnalytics package to examine the correlation among all 12 of these variables.** 

```{r}
#First metric - total area
patchArea = lsm_p_area(forest, directions=8)

patchArea

#Second metric - core area
patchCore = lsm_p_core(forest, directions=8)

patchCore

#Third metric - core area index
patchCai = lsm_p_cai(forest, directions=8)

patchCai

#Fourth metric - euclidean nearest neighbor distance
patchEnn = lsm_p_enn(forest, directions=8)

patchEnn

#Fifth metric - contiguity index
patchCont = lsm_p_contig(forest, directions=8)

patchCont

#Sixth metric - perimeter 
patchPerim = lsm_p_perim(forest, directions=8)

patchPerim

#Seventh metric - perimeter to area ratio 
patchPara = lsm_p_para(forest, directions=8)

patchPara

#Eighth metric - related circumscribing circle 

patchCircle = lsm_p_circle(forest, directions=8)

patchCircle

#Ninth metric - fractal dimension index
 patchFrac = lsm_p_frac(forest, directions=8)
 
 patchFrac
 
#Tenth metric - radius of gyration  
  patchGy = lsm_p_gyrate(forest, directions=8)
  
  patchGy
  
# Eleventh metric - shape index
   patchShp = lsm_p_shape(forest, directions=8)
   
   patchShp 
   
#Twelth metric - number of core areas
   patchNcore = lsm_p_ncore(forest, directions=8)
   
   patchNcore
   

```


```{r}
df = rbind(patchArea, patchCore, patchCai, patchEnn, patchCont, patchPerim, patchPara, patchCircle, patchFrac, patchGy, patchShp, patchNcore)

df2 = pivot_wider(df,names_from = metric, values_from = value)
df3 = df2 %>% select(area:ncore)


PerformanceAnalytics::chart.Correlation(df3)

```
*What patterns do you notice, and what do they tell you about the uniqueness or redundancy of the patch-level metrics you calculated?*

Patch values are very tied to characteristics of each patch, which can vary alot. Small patches lack core areas, areas of gyration, and continguity. Many patches have a nearest-neighbor distance of 60 m. Several metrics measure similar characteristics such as core area to core area index and area, perimeter, and area to perimeter ratio. I think all of these metrics are valuable for studying the ecology of particular organisms, but they might not be used in the same analysis due to the redundancy. 

$\color{red}{\text{Your code simply didn't produce the figure I was looking for, but I fixed it above. +3}}$


## Challenge 2 (4 points)

**In our lab, we used the 8-direction or "queen" rule to delineate patches. Using the nlcdSimple raster we created, explore the differences in patch characteristics if you were to use the 4-direction or "rook" rule for delineating patches. Calculate the following class-level metrics for forest cover for both the queen and rook patch delineation rules: number of patches, mean patch size, standard deviation in patch size, nearest-neighbor distance, and total edge length.** 

```{r}
#The functions annoyingly summarize by class instead of category, so we have to merge the category values back in after the summary.

classCats = data.frame('class' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))


#number of patches
patchNQ = lsm_c_np(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchNQ

patchNR = lsm_c_np(nlcdSimple, directions=4)%>% 
  left_join(classCats, by='class')

patchNR

#mean patch size
patchAreaMQ = lsm_c_area_mn(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchAreaMQ

patchAreaMR = lsm_c_area_mn(nlcdSimple, directions=4)%>% 
  left_join(classCats, by='class')

patchAreaMR

#standard deviation in patch size
patchSDQ = lsm_c_area_sd(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchSDQ

patchSDR = lsm_c_area_sd(nlcdSimple, directions=4)%>% 
  left_join(classCats, by='class')

patchSDR

#nearest-neighbor distance
patchEnnQ = lsm_c_enn_cv(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchEnnQ

patchEnnR = lsm_c_enn_cv(nlcdSimple, directions=4)%>% 
  left_join(classCats, by='class')

patchEnnR

#total edge length
patchTeQ = lsm_c_te(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchTeQ

patchTeR = lsm_c_te(nlcdSimple, directions=4)%>% 
  left_join(classCats, by='class')

patchTeR

```
*What patterns do you notice? When might it be appropriate to use one rule vs. the other?*
Applying the queen or rook rule to delineate patches can affect measurements depending on what you are measuring. When looking at the class level metrics for number of patches, mean patch size, standard deviation in patch size, nearest-neighbor distance, and total edge length, only total edge length did not show any difference between using the queen and rook rules. Nearest-neighbor distance and number of patches usually increases when using the rook rule compared to using the queen rule. Standard deviation in patch size and mean patch size usually decreases when using the rook rule compared to using the queen rule. 
I think that using the rook rule better applies to species with that require larger habitats, due to having larger body sizes, limited movement, or needing more area due to diet or territorial behavior, like mountain lions. The queen rule better applies to species with smaller body sizes, better movement, and/or higher population densities, like most passerines.

$\color{red}{\text{Nice work. +4}}$

## Challenge 3 (4 points)


**Using the same zoomed-in study area that we used in the lab, download NLCD raster data for the years 2001 and 2019 and simplify these rasters into 6 landcover categories (wet, developed, barren, forest, grassland, and agriculture). Plot these two rasters.**

```{r}
studyArea = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week4/studyArea.shp')
nlcd2001 = get_nlcd(studyArea, label='Landscape01', year=2001)

# values(nlcd2001) = as.character(values(nlcd2001))

plot(nlcd2001)

```

```{r}
nlcdSimpleA = nlcd2001

plot(nlcdSimpleA)

sum(nlcd2001[nlcd2001==11])

nlcdSimpleA[nlcdSimpleA== 11] = 1 #Wet areas are a 1 now

plot(nlcdSimpleA)



nlcdSimpleA[nlcdSimpleA %in% c(21, 22, 23)] = 2 #All developed areas are 2
nlcdSimpleA[nlcdSimpleA== 52] = 3 #Barren land and shrub/scrub are 3
nlcdSimpleA[nlcdSimpleA %in% c(41, 42, 43)] = 4 #All forest types are 4
nlcdSimpleA[nlcdSimpleA == 71] = 5 #Grassland is 5
nlcdSimpleA[nlcdSimpleA %in% c(81, 82)] = 6 #And agriculture is 6

#Give these numbers category names
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcdSimpleA = categories(nlcdSimpleA, value=tmp)

#And plot the new raster
ggplot(nlcdSimpleA, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))
```


```{r}
studyArea = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week4/studyArea.shp')
nlcd2019 = get_nlcd(studyArea, label='Landscape19', year=2019)

#values(nlcd2019) = as.character(values(nlcd2019))

plot(nlcd2019)
```

```{r}
nlcdSimpleB = nlcd2019
nlcdSimpleB[nlcdSimpleB==11] = 1 #Wet areas are a 1 now
nlcdSimpleB[nlcdSimpleB %in% c(21, 22, 23, 24)] = 2 #All developed areas are 2
nlcdSimpleB[nlcdSimpleB %in% c(31, 52)] = 3 #Barren land and shrub/scrub are 3
nlcdSimpleB[nlcdSimpleB %in% c(41,42,43)] = 4 #All forest types are 4
nlcdSimpleB[nlcdSimpleB == 71] = 5 #Grassland is 5
nlcdSimpleB[nlcdSimpleB %in% c(81,82)] = 6 #And agriculture is 6

#Give these numbers category names
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcdSimpleB = categories(nlcdSimpleB, value=tmp)

#And plot the new raster
ggplot(nlcdSimpleB, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))
```
**What are some of the initial changes you notice between 2001 and 2019?**

The 2019 area map has less forest and more grassland, agriculture, and open space compared to the 2001 area map.  

**Quantify this at the class level by calculating and reporting the changes in (1) the total amount of each land cover type (2) mean patch size for each land cover type, and (3) mean nearest neighbor distance for each cover type between the years 2011 and 2019.** 
```{r}
#Total amount of each landcover type
patchA2011 = lsm_c_ca(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchA2011

patchA2019 = lsm_c_ca(nlcdSimpleB, directions=8)%>% 
  left_join(classCats, by='class')

patchA2019

#mean patch size for each landcover type
patchArea2011 = lsm_c_area_mn(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchArea2011

patchArea2019 = lsm_c_area_mn(nlcdSimpleB, directions=8)%>% 
  left_join(classCats, by='class')

patchArea2019

#mean nearest neighbor distance
patchE2011 = lsm_c_enn_mn(nlcdSimple, directions=8)%>% 
  left_join(classCats, by='class')

patchE2011

patchE2019 = lsm_c_enn_mn(nlcdSimpleB, directions=8)%>% 
  left_join(classCats, by='class')

patchE2019

```
*Give a short description of how you interpret the changes in these values.*
total landcover areas show that the study site lost forest cover between 2011 and 2019, and it gained area in open, grassland, and agriculture. Likewise, the mean patch size for forest decreased, while mean patch sizes for open, grassland, and agriculture increased. This suggests an increase in anthropogenic usage. Forests may have been lost through logging or natural disturbance. This space was then replaced with landcover that grows back quickly. 

Mean nearest-neighbor distance increased for open, forest, and agriculture. It only decreased for grassland. Along with mean patch size, this means that open, grassland, and agriculture are larger patches. The increase in grassland nearest-neighbor distance could be due to restoration efforts or simply grasses filling suitable open spaces.


**Quantify these changes at the landscape level by calculating and reporting on changes in the (1) Shannon diversity and (2) Shannon evenness of the landscapes at the different time points.**
```{r}
#2001 
lsm_l_shdi(nlcdSimpleA)

lsm_l_shei(nlcdSimpleA)

#2011
lsm_l_shdi(nlcdSimple)

lsm_l_shei(nlcdSimple)

#2019
lsm_l_shdi(nlcdSimpleB)

lsm_l_shei(nlcdSimpleB)

```
*Give a short description of how you interpret the changes in these values.*

Shannon diversity decreased a little bit between 2001 and 2011 but then increased between 2011 and 2019. This means that the landscape diversity decreased between 2001 and 2011 but then increased between 2011 and 2019.  Shannon eveness increased between 2001 and 2011, but decreased between 2011 and 2019. This means that landcover type distribution became more similar over the area from 2001 to 2011 and less similar from 2011 to 2019. Changes from 2001 to 2011 are likely due to increased forest cover and distribution, while changes from 2011 to 2019 could be due to natural disturbance or logging that breaks up forest cover, but mostly removes particular patches.

$\color{red}{\text{Great! +4}}$

## Challenge 4 (4 points)

**Use the voss2d() function to simulate a surface where g = 7 and H = 0.5. From that, create 9 'landscapes' with 10%, 20%, 30%, ..., 90% threshold values. The '1' values here can represent anything your imagination comes up with. It could be forest cover, cover of some other land cover type, bodies of water, temperatures above a threshold, etc. I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes.**

```{r}
set.seed(23)
vossModel = voss2d(g=7, H=0.5)
vossModel = rast(vossModel$z)

plot(vossModel)


```



```{r}
# 90% cover
threshold90 = quantile(as.matrix(vossModel), prob=0.9)
voss90 = ifel(vossModel > threshold90, 0, 1)
plot(voss90)

# 80% cover
threshold80 = quantile(as.matrix(vossModel), prob=0.8)
voss80 = ifel(vossModel > threshold80, 0, 1)
plot(voss80)

# 70% cover
threshold70 = quantile(as.matrix(vossModel), prob=0.7)
voss70 = ifel(vossModel > threshold70, 0, 1)
plot(voss70)

# 60% cover
threshold60 = quantile(as.matrix(vossModel), prob=0.6)
voss60 = ifel(vossModel > threshold60, 0, 1)
plot(voss60)

# 50% cover
threshold50 = quantile(as.matrix(vossModel), prob=0.5)
voss50 = ifel(vossModel > threshold50, 0, 1)
plot(voss50)

# 40% cover
threshold40 = quantile(as.matrix(vossModel), prob=0.4)
voss40 = ifel(vossModel > threshold40, 0, 1)
plot(voss40)

# 30% cover
threshold30 = quantile(as.matrix(vossModel), prob=0.3)
voss30 = ifel(vossModel > threshold30, 0, 1)
plot(voss30)

# 20% cover
threshold20 = quantile(as.matrix(vossModel), prob=0.2)
voss20 = ifel(vossModel > threshold20, 0, 1)
plot(voss20)

# 10% cover
threshold10 = quantile(as.matrix(vossModel), prob=0.1)
voss10 = ifel(vossModel > threshold10, 0, 1)
plot(voss10)

```
*comment on what patterns you see changing as the value increases from a 10% cover to 90% cover*

When comparing the patterns to the model, the landcover starts from the lowest values and expanding into increasing values as the percentage increases. This means that the lowest values are most favorable for presence of the landcover type.  

**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is threshold value and y-axis is calculated metric). .**
```{r}
#90% 
ai90 = lsm_c_ai(voss90)
ai90

coremn90 = lsm_l_core_mn(voss90)
coremn90

edge90 = lsm_l_te(voss90)
edge90

#80%
ai80 = lsm_c_ai(voss80)
ai80

coremn80 = lsm_l_core_mn(voss80)
coremn80

edge80 = lsm_l_te(voss80)
edge80

#70%
ai70 = lsm_c_ai(voss70)
ai70

coremn70 = lsm_l_core_mn(voss70)
coremn70

edge70 = lsm_l_te(voss70)
edge70

#60%
ai60 = lsm_c_ai(voss60)
ai60

coremn60 = lsm_l_core_mn(voss60)
coremn60

edge60 = lsm_l_te(voss60)
edge60

#50%
ai50 = lsm_c_ai(voss50)
ai50

coremn50 = lsm_l_core_mn(voss50)
coremn50

edge50 = lsm_l_te(voss50)
edge50

#40%
ai40 = lsm_c_ai(voss40)
ai40

coremn40 = lsm_l_core_mn(voss40)
coremn40

edge40 = lsm_l_te(voss40)
edge40

#30%
ai30 = lsm_c_ai(voss30)
ai30

coremn30 = lsm_l_core_mn(voss30)
coremn30

edge30 = lsm_l_te(voss30)
edge30

#20%
ai20 = lsm_c_ai(voss20)
ai20

coremn20 = lsm_l_core_mn(voss20)
coremn20

edge20 = lsm_l_te(voss20)
edge20

#10%
ai10 = lsm_c_ai(voss10)
ai10

coremn10 = lsm_l_core_mn(voss10)
coremn10

edge10 = lsm_l_te(voss10)
edge10


```
```{r}
Scaterpl = read.csv('https://raw.githubusercontent.com/GLuchauer/WILD-7970-Labs/main/Part4.csv')
head(Scaterpl)

#plot(Y~X,data=datum4)
#results=lm(Y~X,data=datum4)
#abline(results)
plot(Aggregate_Index~Percent,data=Scaterpl)
plot(Mean_Core_Area~Percent,data=Scaterpl)
plot(Total_Landscape_Edge~Percent,data=Scaterpl)
```
*Briefly describe why you chose these 3 metrics and how they change with increasing cover*

I chose class aggregation index, landscape mean core area, and total landscape edge to show changes in the model with increased landcover. Class 1 Aggregation increases while class 0 decreases with increased percent landcover. I picked aggregation index in particular to show transformation from one landcover type to another. I picked landscape mean core area because I thought it would increase with increased percent landcover. However mean core area often decreased until the higher percent landcovers (80%+). This is likely due to the increasing number of new small core areas. I picked total edge because I thought it would increase and then decrease at higher percent landcover. It is at its highest at 50% cover and decreases with both increasing and decreasing percent landcover.

$\color{red}{\text{Awesome. +4}}$


## Challenge 5 (4 points)

**Use the voss2d() function to simulate 9 surfaces. In each one, g should be 7, but vary the value of H from 0.1 to 0.9. Create a unique landscape from each of these with a threshold value of 30% cover. Again, the 1 values can represent anything you desire, and again I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes.**

```{r}
set.seed(23)
#90%
vossModel9 = voss2d(g=7, H=0.9)
vossModel9 = rast(vossModel9$z)

plot(vossModel9)

threshold309 = quantile(as.matrix(vossModel9), prob=0.3)
H309 = ifel(vossModel9 > threshold309, 0, 1)
plot(H309)

#80%
vossModel8 = voss2d(g=7, H=0.8)
vossModel8 = rast(vossModel8$z)

plot(vossModel8)

threshold308 = quantile(as.matrix(vossModel8), prob=0.3)
H308 = ifel(vossModel8 > threshold308, 0, 1)
plot(H308)

#70%
vossModel7 = voss2d(g=7, H=0.7)
vossModel7 = rast(vossModel7$z)

plot(vossModel7)

threshold307 = quantile(as.matrix(vossModel7), prob=0.3)
H307 = ifel(vossModel7 > threshold307, 0, 1)
plot(H307)

#60%
vossModel6 = voss2d(g=7, H=0.6)
vossModel6 = rast(vossModel6$z)

plot(vossModel6)

threshold306 = quantile(as.matrix(vossModel6), prob=0.3)
H306 = ifel(vossModel6 > threshold306, 0, 1)
plot(H306)

#50%
vossModel5 = voss2d(g=7, H=0.5)
vossModel5 = rast(vossModel5$z)

plot(vossModel5)

threshold305 = quantile(as.matrix(vossModel5), prob=0.3)
H305 = ifel(vossModel5 > threshold305, 0, 1)
plot(H305)

#40%
vossModel4 = voss2d(g=7, H=0.4)
vossModel4 = rast(vossModel4$z)

plot(vossModel4)

threshold304 = quantile(as.matrix(vossModel4), prob=0.3)
H304 = ifel(vossModel4 > threshold304, 0, 1)
plot(H304)

#30% 
vossModel3 = voss2d(g=7, H=0.3)
vossModel3 = rast(vossModel3$z)

plot(vossModel3)

threshold303 = quantile(as.matrix(vossModel3), prob=0.3)
H303 = ifel(vossModel3 > threshold303, 0, 1)
plot(H303)

#20%
vossModel2 = voss2d(g=7, H=0.2)
vossModel2 = rast(vossModel2$z)

plot(vossModel2)

threshold302 = quantile(as.matrix(vossModel2), prob=0.3)
H302 = ifel(vossModel2 > threshold302, 0, 1)
plot(H302)

#10%
vossModel1 = voss2d(g=7, H=0.1)
vossModel1 = rast(vossModel1$z)

plot(vossModel1)

threshold301 = quantile(as.matrix(vossModel1), prob=0.3)
H301 = ifel(vossModel1 > threshold301, 0, 1)
plot(H301)

```
 *comment on what patterns you see changing as the H value increases from 0.1 to 0.9*
 
 As the value of H increases from 0.1 to 0.9,landcover becomes more clumped and less dispersed within the other landcover type. Patches become more defined, which likely increases core area and decreases Shannon eveness. 
 
**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. THESE MUST BE DIFFERENT THAN THOSE METRICS YOU USED IN CHALLENGE 2. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is H-value and y-axis is calculated metric).**

```{r}
#H = 0.9 
clump9 = lsm_c_clumpy(H309)
clump9

core9 = lsm_l_core_cv(H309)
core9

ndca9 = lsm_l_ndca(H309)
ndca9

#H = 0.8
clump8 = lsm_c_clumpy(H308)
clump8

core8 = lsm_l_core_cv(H308)
core8

ndca8 = lsm_l_ndca(H308)
ndca8

#H = 0.7
clump7 = lsm_c_clumpy(H307)
clump7

core7 = lsm_l_core_cv(H307)
core7

ndca7 = lsm_l_ndca(H307)
ndca7

#H = 0.6
clump6 = lsm_c_clumpy(H306)
clump6

core6 = lsm_l_core_cv(H306)
core6

ndca6 = lsm_l_ndca(H306)
ndca6

#H = 0.5
clump5 = lsm_c_clumpy(H305)
clump5

core5 = lsm_l_core_cv(H305)
core5

ndca5 = lsm_l_ndca(H305)
ndca5

#H = 0.4
clump4 = lsm_c_clumpy(H304)
clump4

core4 = lsm_l_core_cv(H304)
core4

ndca4 = lsm_l_ndca(H304)
ndca4

#H = 0.3
clump3 = lsm_c_clumpy(H303)
clump3

core3 = lsm_l_core_cv(H303)
core3

ndca3 = lsm_l_ndca(H303)
ndca3

#H = 0.2
clump2 = lsm_c_clumpy(H302)
clump2

core2 = lsm_l_core_cv(H302)
core2

ndca2 = lsm_l_ndca(H302)
ndca2

#10%
clump1 = lsm_c_clumpy(H301)
clump1

core1 = lsm_l_core_cv(H301)
core1

ndca1 = lsm_l_ndca(H301)
ndca1
# lsm_l_nd

```
```{r}
Scaterpll = read.csv('https://raw.githubusercontent.com/GLuchauer/WILD-7970-Labs/main/Part5.csv')
head(Scaterpll)

#plot(Y~X,data=datum4)
#results=lm(Y~X,data=datum4)
#abline(results)
plot(Clumpiness~H,data=Scaterpll)
plot(Core_Area~H,data=Scaterpll)
plot(Num_Disjunct_Core_Areas~H,data=Scaterpll)
```
*Briefly describe why you chose these 3 metrics and how they change with increasing cover.*

I chose clumpiness index due to the visual appearance of increasing clumpiness with increasing H in the previous section of this assignment. Generally, the clumpiness increases with increasing H. Landscape total core area was selected because I thought it would increase with increasing H. However, total core area generally decreased. This is likely due to increased edge area compared to increased clumpiness. Number of disjunct core areas was selected due because I expect core areas to decrease with increased H due to fewer total patches. This value did decrease, likely due to both fewer patches but also due to less total core area. 

$\color{red}{\text{Great work, Gretchen! +4}}$