---
title: "Lab_6_Spatial"
output: html_document
date: "2024-02-15"
---

```{r}
rm(list=ls())

require(tidyverse)
# require(sf)
require(pgirmess)
require(ncf)
library(geoR)
require(terra)
# library(gstat)
# require(RandomFields)
```

# Background

Researchers set up a square plot with a dimension of 130 m on each side. Within the plot they established blue vane traps in a 5 m grid pattern to sample bee abundance. Every day for 10 days the researchers emptied the blue vane traps and recorded the number of bees collected. After the 10-day sampling period, they calculated the mean daily number of bees captured in the trap each day.

```{r}
blueVanes = read.csv("https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week6/blueVanes.csv")
head(blueVanes)

```



# Challenge 1 (2 points)

Import the blueVanes.csv file from GitHub. Plot the mean bee abundance as a function of the x and y coordinates.

```{r}
ggplot(blueVanes, aes(x=x, y=y, color=meanBee))+
  geom_point()+
  theme_bw()+
  theme(panel.grid=element_blank())+
  scale_color_gradient(low='blue', high='red')
```
*Upon initial inspection, does there appear to be spatial dependence in the bee response data?*

The eyeball test suggests that mean for bees captured daily exhibit some aggregation. Grid shows areas of high mean bee response and areas of low mean bee response. 



# Challenge 2 (3 points)

Empirically test for evidence of spatial dependence in bee abundance by first building a corelogram, then building a semi-variogram. Be sure to provide an envelope for assessing whether the values in each figure are statistically different from what would be expected if the bee values were randomly distributed. 

```{r}
# build matrix
coords = cbind(blueVanes$x, blueVanes$y)
colnames(coords) = c('x', 'y')
distMat = as.matrix(dist(coords))

maxDist = (2/3) * max(distMat)
```

```{r}
#build corelogram
correlogPgirmess = pgirmess::correlog(coords, blueVanes$meanBee, method='Moran', nbclass=14, alternative = 'two.sided')

round(correlogPgirmess, 2)
```

```{r}
#Graphing Correlogram
tmp = data.frame(correlogPgirmess) %>%
  mutate(significant = ifelse(p.value < 0.05, 'yes', 'no'))

ggplot(tmp, aes(x=dist.class, y=coef))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'))+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, 75, 5))


```
semi variogram
```{r}
#Note that this is automatically considering the first two columns to be x and y and the third column to be z.
geoBlue = as.geodata(blueVanes)
plot(geoBlue)
```

```{r}
empGeor = variog(geoBlue, max.dist=maxDist, breaks=seq(0, maxDist, by=3))
plot(empGeor)

emp4Geor = variog4(geoBlue, max.dist=maxDist, direction = c(0, pi/4, pi/2, 3*pi/4))
plot(emp4Geor)
```
*Describe the patterns you see, and your interpretation with regards to the spatial dependence in mean bee abundance. Be sure to identify the maximum distance at which you should interpret your figures (2/3 of the maximum distance between points).*

The correlogram for mean for bees captured daily suggests that the population is not randomly distributed. Correlograms that show random distributions have a slope of zero after the initial close samples with higher likelihood of autocorrelation. This correlogram fluctuates from 25 to 84 which suggests a non-random distribution. The semi-variogram shows a still with a positive slope. This could be due to anisotrophy in the data, which increases with increased distance. If the data were isotropic, then the sill would have a slope of zero. The nugget is likely to be 6.5 to 7, depending on the direction of measurement. 


# Challenge 3 (4 points)

#A
Fit a simple linear model (assuming a normally distributed error term) that models mean bee abundance as a function of an intercept only (no other covariates in the model). 

```{r}
Linear1 = lm(y~x, data = geoBlue)

#mlLinear = likfit(geoBlue, cov.model='linear', ini=c(700, 10))

summary(Linear1)
plot(Linear1)
abline(Linear1)
```

*Compare the model-estimated intercept to the bee abundance values averaged across all sites. What do you notice?*  

The model-estimated intercept for bee abundance is 6.45 bees per square meter at the intercept. However, the plot shows a most sites have a few bees per square meter with a few sites having many bees and a few sites have almost no bees. This suggests a non-random distribution of bees and possible clustering of bees. 

#B
Empirically test for evidence of spatial autocorrelation IN THE MODEL RESIDUALS by first building a correlogram, then building a semi-variogram. Be sure to provide an envelope for assessing whether the values in each figure are statistically different from what would be expected if the bee values were randomly distributed in space. 

```{r}
mlExp = likfit(geoBlue, cov.model='exp', ini=c(700, 10))
summary(mlExp)

mlSph = likfit(geoBlue, cov.model='sph', ini=c(700, 10))
summary(mlSph)

AIC(mlExp, mlSph)
```

```{r}
plot(empGeor)
lines(mlExp, col='blue')
lines(mlSph, col='red')
```

*Compare these figures to those built for the raw data. What similarities or differences to you notice?*

The semi-variogram shows a larger y-intercept than the raw data with the y-intercept being around 8 instead of 6. However the graphs do portray similar trends, given that the raw data is showing residuals. both graphs show increasing positive slopes that reach a near zero slopes. The semi-variogram shows more less of a strong slope trend around a distance of 60, where semivariance decreases a bit with increasing distance. Additional differences include the scales, where the raw data is collecting the residuals given quartile distribution compared to semivariance given distance for the semi-variogram. 


# Challenge 4 (4 points)

Conventional wisdom indicates that the distribution of floral resources may substantially influence bee abundance. In addition, I am  completely making up the following fact: the average bee has a perceptual range of about 15 m. Below I have provided code for importing a raster that shows the number of flowers counted in 1 m grid cells across the entire study area. Use the focal() function from the previous lab to generate a new raster in which each cell value represents the mean number of flowers in cells within a 15 m radius. Plot the distribution of the blue vane traps over the 15 m focal raster. No text required here, just show me the plot.

```{r}
floral = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week6/floralResources.tif')
plot(floral)

probMatrix = focalMat(floral, 15, type='circle', fillNA=FALSE)
probMatrix

```
```{r}
floralOrginal = floral
floralOrginal[is.na(floralOrginal)] = 0

for15m = focal(floralOrginal, probMatrix, fun='sum')
for15m

plot(floralOrginal)
plot(for15m)
```


# Challenge 5 (4 points)

Extract the focal flower value from the raster you just created at each point using the extract() function. You now have an estimate for the average number of flowers per meter around each blue vane trap. Fit a second linear model, but this time include average floral abundance within 15 m as a covariate. Export the residuals from this fitted model, and again, examine a corelogram and variogram of the residuals. As always, be sure to plot random envelopes.

```{r}
tmp1 = extract(for15m, blueVanes[,1:2])

Linear2 = lm(focal_sum ~ID, data = tmp1)
summary(Linear2)
plot(Linear2)
abline(Linear2)

# build matrix
coords = cbind(tmp1$ID, tmp1$focal_sum)
colnames(coords) = c('ID', 'focal_sum')
distMat = as.matrix(dist(coords))

maxDist = (2/3) * max(distMat)


```
```{r}
#build corelogram
correlogPgirmess1 = pgirmess::correlog(coords, tmp1$focal_sum, method='Moran', nbclass=14, alternative = 'two.sided')

round(correlogPgirmess1, 2)
```
```{r}
#Graphing Correlogram
tmp = data.frame(correlogPgirmess1) %>%
  mutate(significant = ifelse(p.value < 0.05, 'yes', 'no'))

ggplot(tmp, aes(x=dist.class, y=coef))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'))+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, 75, 5))


```
```{r}
#Note that this is automatically considering the first two columns to be x and y and the third column to be z.
geoBlue1 = as.geodata(blueVanes)
plot(geoBlue1)

empGeor1 = variog(geoBlue1, max.dist=maxDist, breaks=seq(0, maxDist, by=3))
plot(empGeor1)
```
*Now what do you notice?*
 The residuals have varying fluctuations with the residuals. Residuals vs fitted and the scale-location shows fluctuations in distribution, but the residuals vs leverage has increasing fluctuations in distributions with increasing leverage.This suggests a non-random distribution. The correlogram and semivariogram support this assessment with correlogram having a negative and then positive slope and the semivariogram has a mostly positive slope with distance. These graphs show greater slopes than the bee abundance directly from the raw data. 

# Challenge 6 (3 points)

*What is your assessment regarding the distribution of bee abundance in this study area? Does it exhibit spatial dependence, and if so, what is driving that pattern?*

My assessment of bee abundance in the study area is that they seem to cluster are particular locations. These sites are likely areas that contain hive and bee colonies. Queen bees likely selected these sites for hive due to containing quality habitat for this species of bee. Parameters that would make a site a good place to cluster would include assess to flowers and water, access to structure to place a hive that can be protected from predators, and distance from other hives. Bee colonies would be more successful if they are spaced out from each other in order to minimize resource competition from other colonies of the same species. 

